---
title: "Plant Phenology and GDDs"
author: "Karlee Bradley"
date: "6/26/2019"
output: html_document
---

This is an RMarkdown file that contains all of the code together.

## Plant Phenology Data

#### Here I read in the different packages that I will use.

```{r library}

library(ggplot2)
library(dplyr)
library(lubridate)
library(neonUtilities)
library(devtools)
library(geoNEON)
library(gridExtra)
library(tidyverse)
library(gtools)
library(httr)
library(jsonlite)
library(RColorBrewer)

# set working directory
setwd("~/Documents/data-for-proj")

# setting file paths for where the data will be and where my repo is 
#if (file.exists(
  #'C:/Users/kbradlee')){
  myPathToGit <- "C:/Users/kbradlee/Desktop/data-visualization"
  myPathToData <- "/Users/kbradlee/Desktop/data-visualization"
#}

```

#### Here, I pull the Plant Phenology Data from the NEON API and read it. I read in all of the phenology data rather than just doing it one site at a time.

The zipsByProduct and stackByTable has to be uncommented, and this chunk has to be run by itself. Then comment again and can knit.
```{r pull-data}

# pull plant phenology data via api instead of on local directory
dpid <- as.character('DP1.10055.001')
#zipsByProduct(dpID=dpid, site="all", package="basic")
#stackByTable(paste0(getwd(), "/filesToStack10055"), folder=T)


# read in the data
ind <- read.csv(paste(myPathToData, "filesToStack10055/stackedFiles/phe_perindividual.csv", sep = "/"),  header = T)
status <- read.csv(paste(myPathToData, "filesToStack10055/stackedFiles/phe_statusintensity.csv", sep = "/"), header = T)


```

#### Here I begin to clean the data.

```{r clean-data}

# remove the uid columns for both the ind and the status data
ind <- select(ind, -uid)
status <- select(status, -uid)

# remove duplicate rows in the ind data and the status data
ind_noD <- distinct(ind)
status_noD <- distinct(status)

# rename variables in the status object to have "Stat" at the end of the variable name before joining the datasets
status_noD <- rename(status_noD, editedDateStat = editedDate, measuredByStat = measuredBy, recordedByStat = recordedBy,
                   samplingProtocolVersionStat = samplingProtocolVersion, remarksStat = remarks, dataQFStat = dataQF)

# rename variables in ind so that it doesn't conflict later on when joining the dataframes
ind_noD <- rename(ind_noD, addDate = date)

# convert the date columns from character class to date class for ind and status
ind_noD$editedDate <- as.Date(ind_noD$editedDate)
status_noD$date <- as.Date(status_noD$date)

# retain only the latest editedDate for each individualID on ind and get rid of duplicate dates
ind_lastnoD <- ind_noD %>%
  group_by(individualID) %>%
  filter(editedDate == max(editedDate)) %>%
  group_by(editedDate, individualID) %>%
  filter(row_number() == 1)

```

#### Join the two data frames together into one dataframe that contains the phenology data for all the sites.

```{r join-data}

# join the two dataframes together into one table
# this will be the table that is used to then narrow sites and species and phenophases
phe_ind <- left_join(status_noD, ind_lastnoD)

# removing columns with mostly NAs that aren't needed
phe_ind <- select(phe_ind, -dayOfYear, -samplingProtocolVersionStat, -remarksStat, -dataQFStat)
phe_ind <- select(phe_ind, -geodeticDatum, -coordinateUncertainty, -elevationUncertainty, -elevationUncertainty, -sampleLatitude, 
                  -sampleLongitude, -sampleCoordinateUncertainty, -sampleElevation, -sampleElevationUncertainty, 
                  -identificationQualifier, -vstTag, -dataQF)
# removing other columns that don't seem to be relevant
phe_ind <- select(phe_ind, -measuredByStat, -recordedByStat, -sampleGeodeticDatum, -samplingProtocolVersion,
                  -measuredBy, -identifiedBy, -recordedBy)

# adding other date columns for better date analysis
phe_ind$dayOfYear <- yday(phe_ind$date)
phe_ind$year <- substr(phe_ind$date, 1, 4)
phe_ind$monthDay <- format(phe_ind$date, format = "%m-%d")

# using geoNEON package to find locations
# adding latitude and longitude from geoNEON to phe_ind, and elevation
spatialOnly <- def.extr.geo.os(phe_ind, 'namedLocation', locOnly = T)
phe_ind$latitude <- spatialOnly$api.decimalLatitude[match(phe_ind$namedLocation, spatialOnly$data.locationName)]
phe_ind$longitude <- spatialOnly$api.decimalLongitude[match(phe_ind$namedLocation, spatialOnly$data.locationName)]
phe_ind$elevation <- spatialOnly$api.elevation[match(phe_ind$namedLocation, spatialOnly$data.locationName)]

```

## AGDDS

Here I load the temperature data for a single site and calculate AGDDs with it.


#### Pulling the temperature data from the NEON API and reading it.

This pulls the temperature data for all of the sites specified.

```{r pull-temp-data}

# pulling the air temp data via api
dpid <- as.character('DP1.00002.001')  ##single aspirated air temperature
sites <- c("HARV", "SERC", "UNDE", "UKFS", "ORNL", "CLBJ", "ABBY", "TOOL", "BONA")


#zipsByProduct(dpID=dpid, site=sites, package="basic", avg = "30")
#stackByTable(paste0(myPathToData, "/filesToStack00002"), folder=T)



# load the temp data
SAAT <- read.csv(paste(myPathToData, "filesToStack00002/stackedFiles/SAAT_30min.csv", sep="/"), 
                 stringsAsFactors = F, header = T)
df <- SAAT

```

#### Formatting the data.

```{r format-data}

# define function to convert temp c to f 
c_to_f <- function(x)  (x * 1.8 + 32)


# remove NAs
df <- filter(df, !is.na(tempSingleMean))

# convert df mean temps from c to f
df$meanTempF <- c_to_f(df$tempSingleMean)


#pull date value from dateTime
# pulls the first 10 indices from endDateTime (yyyy-mm-dd) and creates a new column with just the date
df$date <- substr(df$endDateTime, 1, 10)

#create new dataframe with daily values
# groups by date, and for each date, it takes the max mean and makes that the max, the min mean and makes that the min, and the
# mean mean and makes that the mean
day_temp <- df%>%
  group_by(date, siteID)%>%
  filter(year(date)=="2017") %>% # added the year here so that AGDDs starts from beginning of year
  arrange(date) %>% # had to add this so that the dates were in order
  select(siteID, date, verticalPosition, meanTempF)%>%
  mutate(dayMax=max(meanTempF), dayMin=min(meanTempF), dayMean=mean(meanTempF))%>% 
  select(siteID, date, dayMax, dayMin, dayMean)%>%
  distinct() # only keeps one row for each date with the max, min, and mean


##alt mean, consistent with GDD calculations 
### would be interesting to see how different the accumulation curves looks for true mean vs. simplified mean
# takes the min mean and max mean and uses them to find mean2 by averaging them
day_temp$mean2 <- (day_temp$dayMin + day_temp$dayMax)/2

```

### Calculating Growing Degree Days.

```{r GDDs}

#caluculate daily GDD
# base temp is 50F
# if mean2-50<0, then the number of GDDs is 0
# if mean2-50>0, then it rounds the value to the nearest whole number, and that's the number of GDDs for that date
day_temp$GDD <- ifelse(day_temp$mean2-50 < 0, 0, round(day_temp$mean2-50, 0))


#function to add daily GDD values
sumr.2 <- function(x) {
  sapply(1:length(x), function(i) sum(x[1:i]))
}

```


### Calculating AGDDs using the mean temperatures.

```{r AGDD-mean}

#calculate Accumlated GDD
# uses the sumr.2 function to add the GDDs together and puts it in AGDD
# uses the sumr.2 function to add the GDDs for each site
day_temp <- day_temp %>%
  group_by(siteID) %>%
  mutate(AGDD = sumr.2(x=GDD))

day_temp <- ungroup(day_temp)

# adding useful columns to dataframe for date analysis
day_temp$dayOfYear <- yday(day_temp$date)
day_temp$year <- substr(day_temp$date, 1, 4)
day_temp$monthDay <- format(day_temp$date, format = "%m-%d")

```

### Calculating AGDDs using the minimum and maximum temperatures.

Wondered how much the AGDDs would change if calculated this way. This follows the same code as before for a different dataframe.

```{r AGDD-min-max}

df_2 <- filter(SAAT, !is.na(tempSingleMinimum), !is.na(tempSingleMaximum))

# convert df min temps and max temps from c to f
df_2$minTempF <- c_to_f(df_2$tempSingleMinimum)
df_2$maxTempF <- c_to_f(df_2$tempSingleMaximum)

#pull date value from dateTime
# pulls the first 10 indices from endDateTime (yyyy-mm-dd) and creates a new column with just the date
df_2$date <- substr(df_2$endDateTime, 1, 10)


#create new dataframe with daily values
# groups by date, and for each date, it takes the max mean and makes that the max, the min mean and makes that the min, and the
# mean mean and makes that the mean
day_temp_2 <- df_2%>%
  group_by(date, siteID)%>%
  filter(year(date)=="2017") %>% # added the year here so that AGDDs starts from beginning of year
  arrange(date) %>% # had to add this so that the dates were in order
  select(siteID, date, verticalPosition, minTempF, maxTempF) %>%
  mutate(dayMax=max(maxTempF), dayMin=min(minTempF)) %>% 
  select(siteID, date, dayMax, dayMin)%>%
  distinct() # only keeps one row for each date with the max, min, and mean


# calculating the mean for max and mins of each day
day_temp_2$dayMean <- (day_temp_2$dayMax + day_temp_2$dayMin)/2


#caluculate daily GDD
# base temp is 50F
# if mean2-50<0, then the number of GDDs is 0
# if mean2-50>0, then it rounds the value to the nearest whole number, and that's the number of GDDs for that date
day_temp_2$GDD <- ifelse(day_temp_2$dayMean-50 < 0, 0, round(day_temp_2$dayMean-50, 0))

#calculate Accumlated GDD
# uses the sumr.2 function to add the GDDs together and puts it in AGDD
day_temp_2 <- day_temp_2 %>%
  group_by(siteID) %>%
  mutate(AGDD = sumr.2(x=GDD))

day_temp_2 <- ungroup(day_temp_2)

# adding useful columns to dataframe for date analysis
day_temp_2$dayOfYear <- yday(day_temp_2$date)
day_temp_2$year <- substr(day_temp_2$date, 1, 4)
day_temp_2$monthDay <- format(day_temp_2$date, format = "%m-%d")

```

#### AGDD Plots

Plotting AGDDs for all the sites in 2017 using the first method of calculating AGDDs from the mean temperatures.

```{r AGDD-mean-plot}

# plotting AGDDs for 2017 
AGDD_2017 <- ggplot(data=day_temp, aes(x=dayOfYear, y=AGDD,  color=siteID)) +
  geom_path() + xlab("Day of Year") + ggtitle("Aggregated Growing Degree Days in 2017") +
  scale_color_brewer(palette = "Paired")
AGDD_2017

```

Plotting AGDDs for 2017 using the second method of calculating AGDDs from the minimum and maximum temperatures.
Comparing the two different ways to calculate the AGDDS, it looks like there are more AGDDS using the second calculation.

```{r AGDD-min-max-plot}

AGDD_2017_2 <- ggplot(data=day_temp_2, aes(x=dayOfYear, y=AGDD, color = siteID)) +
  geom_path() + xlab("Day of Year") + ggtitle("Aggregated Growing Degree Days in 2017") +
  scale_color_brewer(palette = "Paired")
AGDD_2017_2

```


## All Sites

#### Filtering all of phe_ind for all sites.

```{r dbl-yes-all}

growthFormOfInterest <- "Deciduous broadleaf"
phe_ind <- filter(phe_ind, growthForm %in% growthFormOfInterest)
  
phenophaseOfInterest <- "Leaves"
phe_ind <- filter(phe_ind, phenophaseName %in% phenophaseOfInterest)

phe_ind <- filter(phe_ind, siteID %in% sites)

# look at the total individuals in leaves status by day
sampSize <- count(phe_ind, date)
inStat <- phe_ind %>%
  group_by(date, taxonID, siteID) %>%
  count(phenophaseStatus) 
inStat <- full_join(sampSize, inStat, by = "date")

ungroup(inStat)

# only look at the yes's
inStat_T <- filter(inStat, phenophaseStatus %in% "yes")

```

#### Total Individuals in Leaf

```{r individuals-all}

# plot of the number of individuals in the leaves phenophase
phenoPlot_all <- ggplot(inStat_T, aes(date, n.y, color = siteID)) +
  geom_bar(stat = "identity") +
  ggtitle("Total Individuals in Leaf") +
  xlab("Date") + ylab("Number of Individuals") +
  theme(plot.title = element_text(lineheight = .8, face = "bold", size = 20)) +
  theme(text = element_text(size = 18)) +
  scale_color_brewer(palette = "Set1")
phenoPlot_all

```

#### Percentage in Leaf

```{r percentage-all}

# plot the percentage of individuals in the leaves phenophase
# convert to percentage
inStat_T$percent <- ((inStat_T$n.y)/inStat_T$n.x)*100

# plot
phenoPlot_P_all <- ggplot(inStat_T, aes(date, percent, color = siteID)) +
  geom_bar(stat = "identity") +
  ggtitle("Percentage in the Leaves Phenophase") +
  xlab("Date") + ylab("% of Individuals") +
  theme(plot.title = element_text(lineheight = .8, face = "bold", size = 20)) +
  theme(text = element_text(size = 18))
phenoPlot_P_all

```

#### Percentage in Leaf in 2017

```{r 2017-p-all}

# plot percentage in just 2017
startTime <- as.Date("2017-01-01")
endTime <- as.Date("2017-12-31")
start.end <- c(startTime, endTime)
# plot
leaves17_all <- ggplot(inStat_T, aes(date, percent, color = siteID)) +
  geom_bar(stat = "identity", na.rm = TRUE) +
  ggtitle("Total Individuals in Leaf in 2017") +
  xlab("Date") + ylab("% of Individuals") +
  (scale_x_date(limits = start.end, date_breaks = "1 month", date_labels = "%b")) +
  theme(plot.title = element_text(lineheight = .8, face = "bold", size = 20)) +
  theme(text = element_text(size = 18)) +
  scale_color_brewer(palette = "Set1")
leaves17_all

```

#### Percentage in Leaf in 2016

Not all of the sites have phenology data from 2016.

```{r 2016-p-all}

# plot percentage in just 2016
startTime16 <- as.Date("2016-01-01")
endTime16 <- as.Date("2016-12-31")
start16.end16 <- c(startTime16, endTime16)
#plot
leaves16_all <- ggplot(inStat_T, aes(date, percent, color = siteID)) +
  geom_bar(stat = "identity", na.rm = TRUE) +
  ggtitle("Total Individuals in Leaf in 2016") +
  xlab("Date") + ylab("% of Individuals") +
  (scale_x_date(limits = start16.end16, date_breaks = "1 month", date_labels = "%b")) +
  theme(plot.title = element_text(lineheight = .8, face = "bold", size = 20)) +
  theme(text = element_text(size = 18)) + 
  scale_color_brewer(palette = "Set1")
leaves16_all

```

The graphs are hard to distinguish between sites.



## HARV 

Choosing to work with just the HARV site and analyzing the DBL species in the Leaves phenophase.
How to do the same things with all of the sites without having all of this code 9 other times?...

#### Selecting HARV

```{r harv}

# select site of interest
siteOfInterest <- "HARV"
harv <- filter(inStat_T, siteID %in% siteOfInterest)

```




### HARV plots

#### Total Individuals in Leaf

```{r individuals-harv}

# plot of the number of individuals in the leaves phenophase
phenoPlot_harv <- ggplot(harv, aes(date, n.y)) +
  geom_bar(stat = "identity", na.rm = TRUE) +
  ggtitle("Total Individuals in Leaf") +
  xlab("Date") + ylab("Number of Individuals") +
  theme(plot.title = element_text(lineheight = .8, face = "bold", size = 20)) +
  theme(text = element_text(size = 18))
phenoPlot_harv

```

#### Percentage in Leaf

```{r percentage-harv}

# plot
phenoPlot_P_harv <- ggplot(harv, aes(date, percent)) +
  geom_bar(stat = "identity", na.rm = TRUE) +
  ggtitle("Percentage in the Leaves Phenophase") +
  xlab("Date") + ylab("% of Individuals") +
  theme(plot.title = element_text(lineheight = .8, face = "bold", size = 20)) +
  theme(text = element_text(size = 18))
phenoPlot_P_harv

```

#### Percentage in Leaf in 2017

```{r 2017-p-harv}

# plot percentage in just 2017
startTime <- as.Date("2017-01-01")
endTime <- as.Date("2017-12-31")
start.end <- c(startTime, endTime)
# plot
leaves17_harv <- ggplot(harv, aes(date, percent)) +
  geom_bar(stat = "identity", na.rm = TRUE) +
  ggtitle("Total Individuals in Leaf in 2017") +
  xlab("Date") + ylab("% of Individuals") +
  (scale_x_date(limits = start.end, date_breaks = "1 month", date_labels = "%b")) +
  theme(plot.title = element_text(lineheight = .8, face = "bold", size = 20)) +
  theme(text = element_text(size = 18))
leaves17_harv

```

#### Percentage in Leaf in 2016

```{r 2016-p-harv}

# plot percentage in just 2016
startTime16 <- as.Date("2016-01-01")
endTime16 <- as.Date("2016-12-31")
start16.end16 <- c(startTime16, endTime16)
#plot
leaves16_harv <- ggplot(harv, aes(date, percent)) +
  geom_bar(stat = "identity", na.rm = TRUE) +
  ggtitle("Total Individuals in Leaf in 2016") +
  xlab("Date") + ylab("% of Individuals") +
  (scale_x_date(limits = start16.end16, date_breaks = "1 month", date_labels = "%b")) +
  theme(plot.title = element_text(lineheight = .8, face = "bold", size = 20)) +
  theme(text = element_text(size = 18))
leaves16_harv

```

#### Comparing Percentage in Leaf in 2016 and 2017

```{r p-16-17-harv}

# plot % leaves in 2016 and 2017 together
grid.arrange(leaves16_harv, leaves17_harv)

```

#### Making a separate dataframe with only 2017 data

```{r only-2017}

# adding the dayOfYear column to inStat_T so it can be used in graphs
inStat_T$dayOfYear <- yday(inStat_T$date)


# making another dataframe that only has info from 2017
inStat_T_17 <- inStat_T %>%
  filter(year(date)=="2017")

```

#### Percentage of Different Species in 2017

```{r bar-graph}



# bar graph that shows how much of 100% is ACRU and how much is QURU for each day of the year
# can see that for some days, both of them aren't in leaves, ACRU looks like it changes faster than QURU
DBL_P_2017 <- ggplot(data = inStat_T_17, aes(x=dayOfYear, y=percent, group = 1, color = taxonID)) +
  geom_col() + xlab("Day of Year") + ylab("% of DBL") +
  ggtitle("Percent of Deciduous Broadleaf Species in Leaf in 2017") +
  labs(color = "Species") + theme(legend.position = "top") + xlim(c(0,365))
DBL_P_2017

```

## HARV Leaves and AGDDs for 2017

Looking at how the GDDs are accumulated over time in 2017 and how the leaves phenophase changes in 2017.

#### Percentage in Leaf and AGDDs in 2017

```{r 2017-AGGDs}

# putting the % in Leaf and AGDD together
grid.arrange(DBL_P_2017, AGDD_2017)

```


#### Zoomed Plots

```{r zoom}

# zooming on on graph so that it's easier to compare with AGDD_2017
DBL_P_2017 <- ggplot(data = inStat_T_17, aes(x=dayOfYear, y=percent, group = 1, color = taxonID)) +
  geom_col() + xlab("Day of Year") + ylab("% of DBL") +
  ggtitle("Percent of Deciduous Broadleaf Species in Leaf in 2017") +
  labs(color = "Species") + theme(legend.position = "top") + xlim(c(100,350))
DBL_P_2017

# zooming in on the graph so easier to compare with DBL_P_2017
AGDD_2017_ <- ggplot(data=day_temp, aes(x=dayOfYear, y=AGDD, group = 1)) +
  geom_path() + xlab("Day of Year") + ggtitle("Aggregated Growing Degree Days in 2017") + xlim(c(100, 350))
AGDD_2017_

# putting the % in Leaf and AGDD together after zoomed in
grid.arrange(DBL_P_2017, AGDD_2017)

```

#### AGDDs Calculation 2 and DBL Percentage, Zoomed

```{r min-max-zoom}

# zooming in on the graph so easier to compare with DBL_P_2017
AGDD_2017_2_ <- ggplot(data=day_temp_2, aes(x=dayOfYear, y=AGDD, group = 1)) +
  geom_path() + xlab("Day of Year") + ggtitle("Aggregated Growing Degree Days in 2017") + xlim(c(100, 350))
AGDD_2017_2_

# putting the % in Leaf and AGDD_2 together after zoomed in
grid.arrange(DBL_P_2017, AGDD_2017_2)

```





